### 2017-08-28 sec-related-work.tex
Whitespace changes and minor edits.
Address copy-editing feedback.

Our model is based heavily on the prior art of Hidden Markov
Models (HMMs)~\cite{Rabiner:1990:RSR} for sequence labeling tasks.  As a member
of the more general family of probabilistic graphical models~\cite{Koller:2009:PGM}, HMMs are widely applicable and have been used for tasks such as speech
recognition~\cite{Huang:1990:HMM}, part-of-speech
tagging~\cite{Jurafsky:2009:SLP}, and econometrics~\cite{Hamilton:1990:JoE}.
%and are a member of the more general family of probabilistic graphical
%models~\cite{Koller:2009:PGM}.  
A major challenge in applying HMMs and
other graphical models successfully to solve a problem is to design an
appropriate architecture of the model, which always varies according to
specific applications.

For example, in part-of-speech tagging~\cite{Jurafsky:2009:SLP}, the output
distributions are categorical (distributions over words from a fixed
vocabulary) and the latent states represent the part-of-speech category for
a word. In speech recognition~\cite{Huang:1990:HMM}, the output
distributions might be mixtures of Gaussians to predict real-valued vectors
extracted from short windows of a speech signal. In the domain of
econometrics, \citet{Hamilton:1990:JoE} explores HMMs in the context of
``regime-switching.'' In this framing, the goal is to understand how
econometric data changes by modeling discrete changes in ``regime'' as
having an impact on the resulting real-valued vector data observed. The
``regimes'' are represented with some sort of model that can produce
real-valued vector data, such as a multivariate Gaussian or an
auto-regressive model. The analogy with HMMs is that a ``regime'' is a
latent state, and the characterization of the regime itself is the output
distribution for that latent state. Our model can be seen as such a
``regime-switching'' model where the output of the ``regimes'' that
students are switching between are discrete-valued \emph{sequences} (as
opposed to real numbers, vectors of real numbers, or categorical symbols)
and the model used to represent a specific ``regime'' is an (observable)
Markov chain over the observed student actions. We view the switching
between ``regimes'' as the first ``layer'' of our model, and the
transitioning behavior \emph{within} a ``regime'' between the actions a
student takes as the second ``layer'' of our model.

@@

Our formulation differs from both \citet{Zhang:2004:CVPR} and
\citet{Oliver:2004:CVIU} in that we do not feed the labeled sequence of the
lower level into the input of the higher level. Instead, our lower level is
actually treated using a non-hidden Markov model, and the higher level is
modeling transitions between the $K$ different non-hidden Markov models we
consider. The problem to be solved is similar in that we wish to predict a
``label'' for a sequence of actions a student takes as well as understand
the transition behavior between those labels.  However, one of the
consequences of modeling the lower layer using a non-hidden Markov model
instead of an HMM directly is that the meaning of the $K$ different latent
states can be more clearly captured by using our proposed behavior
representation. If we were to use an HMM as our first layer, the behavior
patterns (like from Figure~\ref{fig:motivating-example}) would instead have
nodes that represent another set of latent states instead of being concrete
actions themselves. Each of these latent states would then be associated
with some other output distribution over the possible concrete actions to
be considered. In order to understand a single behavior pattern uncovered,
one would first have to understand the different output distributions for
the latent states in that pattern in order to understand the meaning of
that latent state. This further complicates the understanding of higher
level patterns because understanding the higher layer patterns requires
understanding the lower layer patterns. By taking a more restrictive view
of the first layer, we are able to produce a representation that can be
more readily interpreted due to the states in our first layer
representation having an immediately clear meaning (the concrete action
they represent).

@@

For example, \citet{Kizilcec:2013:LAK} assigned students to states
following a rule-based approach based upon when the student submitted the
assignment for a particular week in the course. They investigated how
students transitioned between these states as the course progressed, and
used the sequence of states a student exhibited as a representation for
performing $k$-means clustering of students into related groups. This
differs from our method substantially: we assign students to states using a
probabilistic framework to account for uncertainty in this state
assignment and jointly learn \emph{representations} for these states,
which are treated as being \emph{latent} as opposed to pre-defined using
some rule (or set of rules).  Furthermore, our model provides more
flexibility in how the time segments are defined, allowing for both finer
(for example, day-by-day) or coarser (for example, month-by-month)
granularity. \citet{Shih:2010:EDM} investigated the use of HMM-based
clustering techniques for automatic discovery of student learning
strategies when solving a particular problem.  This is similar to our
approach in that the description of behavior profiles is a Markov model,
but cannot further characterize each latent state with another informative
HMM. Thus their work can be regarded to modeling ``micro'' behavior,
whereas our model can model both ``micro'' and ``macro'' behavior.

\citet{Davis:2016:EDM} investigate frequent student behavior pattern chains
with a set of actions that is defined similarly to ours. However, their
method for finding the common behavioral patterns involves a manual
clustering step to identify behavioral ``motifs,'' which is then followed
by an automatic (rule-based) assignment of all sequences to these motifs.
Our method, by contrast, attempts to do this automatically: the latent
state representations obtained by our model attempt to capture similar
meanings to their behavior motifs in a completely automated fashion. They
also automatically generate and investigate Markov models for different
MOOCs, but do so by considering \emph{all} student action sequences as
belonging to a \emph{single Markov model}. In our approach, we allow each
student behavior sequence to belong to one of $K$ different Markov models
(and further model the transition probabilities between these latent state
Markov models between each sequence a student generates). Thus, their
Markov models presented are a special case of our model when $K = 1$.

>>>>

Our model is based heavily on the prior art of Hidden Markov Models
(HMMs)~\cite{Rabiner:1990:RSR} for sequence labeling tasks. As a member of
the more general family of probabilistic graphical
models~\cite{Koller:2009:PGM}, HMMs are widely applicable and have been
used for tasks such as speech recognition~\cite{Huang:1990:HMM},
part-of-speech tagging~\cite{Jurafsky:2009:SLP}, and
econometrics~\cite{Hamilton:1990:JoE}. A major challenge in applying HMMs
and other graphical models successfully to solve a problem is to design an
appropriate architecture of the model, which always varies according to
specific applications.

For example, in part-of-speech tagging~\cite{Jurafsky:2009:SLP}, the output
distributions are categorical (distributions over words from a fixed
vocabulary) and the latent states represent the part-of-speech category for
a word. In speech recognition~\cite{Huang:1990:HMM}, the output
distributions might be mixtures of Gaussians to predict real-valued vectors
extracted from short windows of a speech signal. In the domain of
econometrics, \citet{Hamilton:1990:JoE} explores HMMs in the context of
``regime-switching.'' In this framing, the goal is to understand how
econometric data changes by modeling discrete changes in ``regime'' as
having an impact on the resulting real-valued vector data observed. The
``regimes'' are thus represented with a model that can produce real-valued
vector data, such as a multivariate Gaussian or an auto-regressive model.
The analogy with HMMs is that a ``regime'' is a latent state, and the
characterization of the regime itself is the output distribution for that
latent state. Our model can be seen as such a ``regime-switching'' model
where the output of the ``regimes'' that students are switching between are
discrete-valued \emph{sequences} (as opposed to real numbers, vectors of
real numbers, or categorical symbols) and the model used to represent a
specific ``regime'' is an (observable) Markov chain over the observed
student actions. We view the switching between ``regimes'' as the first
``layer'' of our model, and the transitioning behavior \emph{within} a
``regime'' between the actions a student takes as the second ``layer'' of
our model.

@@

Our formulation differs from both \citet{Zhang:2004:CVPR} and
\citet{Oliver:2004:CVIU} in that we do not feed the labeled sequence of the
lower level into the input of the higher level. Instead, our lower level is
treated using a non-hidden Markov model, and the higher level is modeling
transitions between the $K$ different non-hidden Markov models we consider.
The problem to be solved is similar in that we wish to predict a ``label''
for a sequence of actions a student takes as well as understand the
transition behavior between those labels.  However, one of the consequences
of modeling the lower layer using a non-hidden Markov model instead of an
HMM directly is that the meaning of the $K$ different latent states can be
more clearly captured by using our proposed behavior representation. If we
were to use an HMM as our first layer, the behavior patterns (like from
Figure~\ref{fig:motivating-example}) would instead have nodes that
represent another set of latent states instead of being concrete actions
themselves. Each of these latent states would then be associated with some
other output distribution over the possible concrete actions to be
considered. In order to understand a single behavior pattern uncovered, one
would first have to understand the different output distributions for the
latent states in that pattern to understand the meaning of that latent
state. This further complicates the understanding of higher level patterns
because understanding the higher layer patterns requires understanding the
lower layer patterns. By taking a more restrictive view of the first layer,
we can produce a representation that can be more readily interpreted due to
the states in our first layer representation having an immediately clear
meaning (the concrete action they represent).

@@

For example, \citet{Kizilcec:2013:LAK} assigned students to states
following a rule-based approach based upon when the student submitted the
assignment for a particular week in the course. They investigated how
students transitioned between these states as the course progressed, and
used the sequence of states a student exhibited as a representation for
performing $k$-means clustering of students into related groups. This
differs from our method substantially: we assign students to states using a
probabilistic framework to account for uncertainty in this state
assignment and jointly learn \emph{representations} for these states,
which are treated as being \emph{latent} as opposed to pre-defined using
some rule (or set of rules).  Furthermore, our model provides more
flexibility in how the time segments are defined, allowing for both finer
(for example, day-by-day) or coarser (for example, month-by-month)
granularity. \citet{Shih:2010:EDM} investigated the use of HMM-based
clustering techniques for automatic discovery of student learning
strategies when solving a problem. This is similar to our approach in that
the description of behavior profiles is a Markov model, but cannot further
characterize each latent state with another informative HMM. Thus, their
work can be regarded to modeling ``micro'' behavior, whereas our model can
model both ``micro'' and ``macro'' behavior.

\citet{Davis:2016:EDM} investigate frequent student behavior pattern chains
with a set of actions that are defined similarly to ours. However, their
method for finding the common behavioral patterns involves a manual
clustering step to identify behavioral ``motifs,'' which is then followed
by an automatic (rule-based) assignment of all sequences to these motifs.
Our method, by contrast, attempts to do this automatically: the latent
state representations obtained by our model attempt to capture similar
meanings to their behavior motifs in a completely automated fashion. They
also automatically generate and investigate Markov models for different
MOOCs, but do so by considering \emph{all} student action sequences as
belonging to a \emph{single Markov model}. In our approach, we allow each
student behavior sequence to belong to one of $K$ different Markov models
(and further model the transition probabilities between these latent state
Markov models between each sequence a student generates). Thus, their
Markov models presented are a special case of our model when $K = 1$.

### 2017-08-28 sec-results.tex
Add link to the source code for the analysis.
Address copy-editing feedback.

\section{Experiment Results}
As an analysis tool, the 2L-HMM model provides us with the following two
patterns to characterize student behavior:
\begin{enumerate*}[label=(\arabic*)]
  \item the latent state representations, and
  \item the latent state transitions.
\end{enumerate*}
Thus to evaluate the proposed model, we conduct experiments to
qualitatively analyze both types of patterns discovered from empirical MOOC
log data.

@@

To visualize these Markov models that represent our latent states, we plot
them as a directed graph where we set the size of a node to be proportional
to its personalized Pagerank score~\cite{Page:1999:PageRank, Jeh:2003:WWW}
where the personalization vector is the initial state distribution for the
Markov model. We let the thickness of a directed edge $(u, v)$ reflect the
probability of taking that edge given that a random walk is currently at
note $u$ (as indicated by the transition matrix)\footnote{We do not plot
the transition probabilities directly within the figure to ease
readability; we instead will mention relevant transition probabilities in
the text as we discuss the plots. The plots were created using
python-igraph: \url{http://igraph.org/python/}.}.

@@

We also argue that it is important that the latent state representation is
a Markov model rather than just a discrete distribution over actions in
$\mathbf{A}$ (as would be the case for a traditional single-layer HMM). We
can observe why if we take a closer look at each of the two latent state
representations in Figure~\ref{fig:states} and look at their forum
component (bottom right). We can see that the relative probability of the
forum activities is roughly the same between these states, but the
\emph{transitions} are quite different. In
Figure~\ref{fig:practice-quiz-state} we have a relatively low probability
of walking from the ``forum thread view'' action back to the ``forum thread
list'' action (see the bottom rightmost arc; transition probability $p
\approx 0.17$), but in Figure~\ref{fig:lecture-viewing-state} we actually
observe a much stronger link in this direction (transition probability $p
\approx 0.63$). This difference highlights an important distinction between
these two latent states: in the first you are more likely to visit the
forum \emph{looking for a particular post}, where in the second you are
more likely to visit the forum \emph{to browse existing posts}. Thus,
capturing the action transition matrix within a latent state is important
for capturing detailed insights involving bigrams of actions.

@@

We can also use our model for cross-course behavior analysis. In
Figure~\ref{fig:cross-course} we show two latent state representations
learned by a 6-state 2L-HMM, one from \textretrieval{} and one from
\sustain{}. These two states were chosen as they are the most similar
between the two courses. However, if we look at the transitions we can see
some important differences. First, in the state from \textretrieval{}, we
can see that the probability of returning to the course wiki after viewing
a lecture (transition probability $p \approx 0.24$) is considerably lower
than that probability in the state from \sustain{} (transition probability
$p \approx 0.57$). We also notice that the self-loop for staying in a
lecture activity in \textretrieval{} (transition probability $p \approx
0.70$) is significantly higher probability than it is in the state from
\sustain{} (transition probability $p \approx 0.34$).  This gives us some
insight into the lecture viewing behavior of students in these two MOOCs
which can possibly be a reflection of the course's structure (which, as
demonstrated in \citet{Davis:2016:EDM}, can differ substantially across
different MOOCs).  In \textretrieval{}, students are likely to view lecture
videos in succession directly without first visiting the course wiki,
whereas in \sustain{} students are much more likely to first return the
course wiki before watching the next lecture video. This observation would
be lost if we did not consider the transitions between the actions within
the latent states.

\subsection{Varying the Number of Latent States}
Our 2L-HMM model has a parameter $K$ that sets the number of latent states
to be learned. We believe that this can allow our model to flexibly
discover patterns of different granularity, and we can show this by varying
$K$ for a particular course and observing how the latent state
representations evolve.

@@

In Figure~\ref{fig:text-state-evolution} we see\footnote{The graphs are
quite small, but the states are in the same positions as they were in
previous figures.} the evolution of the latent state
representations found for the \textretrieval{} MOOC.  With $K=2$ we uncover
a video watching pattern (state 1) and a course material browsing pattern
(state 0). However, we can see that when $K=3$ we are able to uncover a new
pattern involving forum behavior in state 3 (notice the node sizes on the
bottom right). As we increase $K$ to four, we can see that state 1 splits
into state 1 and state 3. These states appear quite similar at a glance,
but there are still some key differences. First, we can see that state 1
now has a non-negligible forum component, whereas state 3 has hardly any
weight on the forum.

@@

In Figure~\ref{fig:trans-low} we plot the latent state transition diagram
for a second group of ``low'' students. These students were selected so
that they attempted all required quizzes in the course, but such that their
average quiz score was $\leq 70\%$. Here, we see that \emph{state 1} has a
large increase in size, where we might have expected state 3 to grow
instead.  However, there is an alternative explanation for this phenomenon.
Since state 1 seems to indicate a highly engaged student, it is a perfectly
reasonable explanation for the ``low'' student group as they are going to
be working hard to try to fill in the gaps in their knowledge. By contrast,
the ``perfect'' student group likely has many members who can take the quiz
more passively and get perfect marks, perhaps because they already know
much of the material being presented, or are just naturally strong and do
not require much background review to perform well. This also explains why
state 1 did not increase in size for the ``perfect'' group like we were
anticipating. \citet{Kizilcec:2017:CandE} observe similar phenomena with
the courses they studied where they find that certificate earning is
negatively correlated with help seeking behavior. Our model lends itself
well to discovering this potentially counter-intuitive insight directly
from data.

@@

Returning to Figure~\ref{fig:trans-comp}, we can also see a difference in
the transitions between the latent states. In particular, look at the
transitions in Figure~\ref{fig:trans-perfect} and
Figure~\ref{fig:trans-low} that leave state 2. In the ``perfect'' group,
nearly all of this probability mass is allocated for the self-loop ($p
\approx 0.93$). In the ``low'' group, this self-loop is less strong ($p
\approx 0.80$; most easily seen by noting that the edges leaving state 2
are darker than for the ``perfect'' group). We can perform a similar
experiment to above by producing a ranked list of students $\ell \in
\mathbf{L}_q$ by their staying probability for state 2 (that is, given that
a student $\ell$ is already in state 2, how likely are they to remain there
in the next action sequence?). This can be computed as
\begin{equation}
  p_\ell(2, 2) = \frac{\sum_{t=1}^{T-1} \xi_t^{(\mathbf{o}_\ell)}(2,2)}
  {\sum_{i=1}^K \sum_{t=1}^{T-1} \xi_t^{(\mathbf{o}_\ell)}(2,i)}
\end{equation}
where $\xi$ is defined as before and computed using the Baum-Welch
algorithm. The last column of Table~\ref{table:mean-rank} indicates that
this transition feature also correlates with the achievement group and
results in significant differences in mean rank.

>>>>

\section{Experiment Results}
As an analysis tool, the 2L-HMM model provides us with the following two
patterns to characterize student behavior:
\begin{enumerate*}[label=(\arabic*)]
  \item the latent state representations, and
  \item the latent state transitions.
\end{enumerate*}
Thus, to evaluate the proposed model, we conduct experiments to
qualitatively analyze both types of patterns discovered from empirical MOOC
log data.

@@

To visualize these Markov models that represent our latent states, we plot
them as a directed graph where we set the size of a node to be proportional
to its personalized Pagerank score~\cite{Page:1999:PageRank, Jeh:2003:WWW}
where the personalization vector is the initial state distribution for the
Markov model. We let the thickness of a directed edge $(u, v)$ reflect the
probability of taking that edge given that a random walk is currently at
note $u$ (as indicated by the transition matrix)\footnote{We do not plot
the transition probabilities directly within the figure to ease
readability; we instead will mention relevant transition probabilities in
the text as we discuss the plots. The plots were created using
python-igraph: \url{http://igraph.org/python/}.}. In the interest of
reproducibility, the source code for analyzing the MOOC logs we use in this
paper and for producing the figures themselves is publicly available
as open-source
software\footnote{\url{https://github.com/skystrife/clickstream-hmm}}.

@@

We also argue that it is important that the latent state representation is
a Markov model rather than just a discrete distribution over actions in
$\mathbf{A}$ (as would be the case for a traditional single-layer HMM). We
can observe why if we take a closer look at each of the two latent state
representations in Figure~\ref{fig:states} and look at their forum
component (bottom right). We can see that the relative probability of the
forum activities is roughly the same between these states, but the
\emph{transitions} are quite different. In
Figure~\ref{fig:practice-quiz-state} we have a relatively low probability
of walking from the ``forum thread view'' action back to the ``forum thread
list'' action (see the bottom rightmost arc; transition probability $p
\approx 0.17$), but in Figure~\ref{fig:lecture-viewing-state} we actually
observe a much stronger link in this direction (transition probability $p
\approx 0.63$). This difference highlights an important distinction between
these two latent states: in the first you are more likely to visit the
forum \emph{looking for a post}, where in the second you are more likely to
visit the forum \emph{to browse existing posts}. Thus, capturing the action
transition matrix within a latent state is important for capturing detailed
insights involving bigrams of actions.

@@

We can also use our model for cross-course behavior analysis. In
Figure~\ref{fig:cross-course} we show two latent state representations
learned by a 6-state 2L-HMM, one from \textretrieval{} and one from
\sustain{}. These two states were chosen as they are the most similar
between the two courses. However, if we look at the transitions we can see
some important differences. First, in the state from \textretrieval{}, we
can see that the probability of returning to the course wiki after viewing
a lecture (transition probability $p \approx 0.24$) is considerably lower
than that probability in the state from \sustain{} (transition probability
$p \approx 0.57$). We also notice that the self-loop for staying in a
lecture activity in \textretrieval{} (transition probability $p \approx
0.70$) is significantly higher probability than it is in the state from
\sustain{} (transition probability $p \approx 0.34$).  This gives us some
insight into the lecture viewing behavior of students in these two MOOCs
which might reflect the course's structure (which, as demonstrated in
\citet{Davis:2016:EDM}, can differ substantially across different MOOCs).
In \textretrieval{}, students are likely to view lecture videos in
succession directly without first visiting the course wiki, whereas in
\sustain{} students are much more likely to first return the course wiki
before watching the next lecture video. This observation would be lost if
we did not consider the transitions between the actions within the latent
states.

\subsection{Varying the Number of Latent States}
Our 2L-HMM model has a parameter $K$ that sets the number of latent states
to be learned. We believe that this can allow our model to flexibly
discover patterns of different granularity, and we can show this by varying
$K$ for a course and observing how the latent state representations evolve.

@@

In Figure~\ref{fig:text-state-evolution} we see\footnote{The graphs are
quite small, but the states are in the same positions as they were in
previous figures.} the evolution of the latent state
representations found for the \textretrieval{} MOOC.  With $K=2$ we uncover
a video watching pattern (state 1) and a course material browsing pattern
(state 0). However, we can see that when $K=3$ we can uncover a new pattern
involving forum behavior in state 3 (notice the node sizes on the bottom
right). As we increase $K$ to four, we can see that state 1 splits into
state 1 and state 3. These states appear quite similar at a glance, but
there are still some key differences. First, we can see that state 1 now
has a non-negligible forum component, whereas state 3 has hardly any weight
on the forum.

@@

In Figure~\ref{fig:trans-low} we plot the latent state transition diagram
for a second group of ``low'' students. These students were selected so
that they attempted all required quizzes in the course, but such that their
average quiz score was $\leq 70\%$. Here, we see that state 1 has a
large increase in size, where we might have expected state 3 to grow
instead.  However, there is an alternative explanation for this phenomenon.
Since state 1 seems to indicate a highly engaged student, it is a perfectly
reasonable explanation for the ``low'' student group as they are going to
be working hard to try to fill in the gaps in their knowledge. By contrast,
the ``perfect'' student group likely has many members who can take the quiz
more passively and get perfect marks, perhaps because they already know
much of the material being presented, or are just naturally strong and do
not require much background review to perform well. This also explains why
state 1 did not increase in size for the ``perfect'' group like we were
anticipating. \citet{Kizilcec:2017:CandE} observe similar phenomena with
the courses they studied where they find that certificate earning is
negatively correlated with help seeking behavior. Our model lends itself
well to discovering this potentially counter-intuitive insight directly
from data.

@@

Returning to Figure~\ref{fig:trans-comp}, we can also see a difference in
the transitions between the latent states. In particular, look at the
transitions in Figure~\ref{fig:trans-perfect} and
Figure~\ref{fig:trans-low} that leave state 2. In the ``perfect'' group,
nearly all this probability mass is allocated for the self-loop ($p
\approx 0.93$). In the ``low'' group, this self-loop is less strong ($p
\approx 0.80$; most easily seen by noting that the edges leaving state 2
are darker than for the ``perfect'' group). We can perform a similar
experiment to above by producing a ranked list of students $\ell \in
\mathbf{L}_q$ by their staying probability for state 2 (that is, given that
a student $\ell$ is already in state 2, how likely are they to remain there
in the next action sequence?). This can be computed as
\begin{equation}
  p_\ell(2, 2) = \frac{\sum_{t=1}^{T-1} \xi_t^{(\mathbf{o}_\ell)}(2,2)}
  {\sum_{i=1}^K \sum_{t=1}^{T-1} \xi_t^{(\mathbf{o}_\ell)}(2,i)}
\end{equation}
where $\xi$ is defined as before and computed using the Baum-Welch
algorithm. The last column of Table~\ref{table:mean-rank} indicates that
this transition feature also correlates with the achievement group and
results in significant differences in mean rank.
